Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-08 Manuscript/TEM_draft.tex
Removed extra comments

We find that, when using a clade conservative metric such as Robinson-Foulds distance, Bayesian consensus trees recovers the right topology better than Maximum Likelihood or than a subset of Bayesian posterior tree distributions and this regardless of the amount of missing data (minimal tree similarity of 0.69 in the worst scenario). Regarding the data, good topological recovery is not related to the amount of missing data \textit{per se} but to the amount of data overlap. The two factors that influences the most this data overlap are the parameters 1 (proportion of coded living taxa) and 3 (number of morphological characters).

%Probably needs to be deeeply rewritten, as said last time to Kev quoting Stuart Pim, people only read the abstract.
% NC: Yep, for now I'll leave this

\end{abstract}

@@

%TG: I completely removed this paragraph
%The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,sansombias2014}. Missing molecular data has been seen by some authors as an issue because it can, in some parts of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing the inference of topology and branch lengths), especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by: (i) including a "modest" number (\texttildelow % NC: This symbol doesn't show up on my PDF. Check.
%50) of highly-covered genes (i.e. a number of genes that are available for all taxa; \citealt{rouresite-specific2011}); (ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and (iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013,pattinsonphylogeny2014}. Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies (e.g. \citealt{sansomfossilization2013}, but see \citealt{pattinsonphylogeny2014}). However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{kearneyfragmentary2002,wiensmissing2003,pattinsonphylogeny2014}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold where it starts fucking up the phylogeny) but more the fact that Bayesian fails.
% NC: OK fair enough. I think though if this is too long we can probably cut this paragraph and put a bit of it into the paragraph below.

@@

\subsection{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the "best" tree topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}).
The Normalised Robinson-Foulds distance is always substantially lower for the Maximum Likelihood %(minimum Normalised Robinson-Foulds distance for $M_{L}$, $M_{F}$ and $M_{C}$ of respectively @, @ and @) %minimum RF value for each parameter individually% TG: or useless (just Bhattacharrya)?
 trees than for the Bayesian consensus trees %(minimum Normalised Robinson-Foulds distance for $M_{L}$, $M_{F}$ and $M_{C}$ of respectively @, @ and @)  %minimum Tr value for each parameter individually% TG: or useless (just Bhattacharrya)?
(median Bhattacharrya Coefficient = @, @, @ %BC value for each parameter individually between both methods RF
$M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). The Normalised Triplets distance is similar between the Maximum Likelihood trees and the Bayesian consensus trees for the $M_{L}$ parameter (median Bhattacharrya Coefficient = @ %BC value for ML between both methods
; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}), however, for the two other parameters, $M_{F}$ and $M_{C}$, the Normalised Triplets distance is substantially higher for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = @, @  %BC value for each parameter individually between both methods Tr
$M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

@@

\section{Acknowledgements}
Thanks to Fr\'{e}d\'{e}ric Delsuc, Emmanuel Douzery, Andrew Jackson, Gavin Thomas and April Wright %TG : Not sure how many people we have to thank but it could also include Graham Slater, Nick Matzke and Trevor Hodkinson.
 % NC: Might be nice to thank Trevor, and it won't add too many lines to thank them all. You can also just calle people A.Wright if you want to reduce word count.
for useful comments on our simulation protocol.
%Thanks to Sive Finlay, Kevin Healy and Adam Kane for comments on the manuscript. %TG: or helping writing sentences does not really count. % NC: Depends how much they helped. I just used to thank the lab for that stuff
Thanks to Paddy Doyle, Graziano D'Innocenzo and Sean McGrath for assistance with the computer cluster. Simulations used the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing and funded through grants from Science Foundation Ireland. This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).

@@

\bibliographystyle{sysbio}
\bibliography{References}
%Needs to be cleaned a bit
% NC: Yeah. Leave it til we send it to Gavin but it should be a priority. Almost all of them are wrong! 
%---------------------------------------------
%
%       Supplementary, separated
%
%---------------------------------------------

>>>>

We find that, when using a clade conservative metric such as Robinson-Foulds distance, Bayesian consensus trees recovers the right topology better than Maximum Likelihood or than a subset of Bayesian posterior tree distributions and this regardless of the amount of missing data (minimal tree similarity of 0.69 in the worst scenario). Regarding the data, good topological recovery is not related to the amount of missing data \textit{per se} but to the amount of data overlap. The two factors that influences the most this data overlap are the parameters 1 (proportion of coded living taxa) and 3 (number of morphological characters).
\end{abstract}

@@

@@

\subsection{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the "best" tree topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}).
The Normalised Robinson-Foulds distance is always substantially lower for the Maximum Likelihood  trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = @, @ and @ %BC value for each parameter individually between both methods RF
for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). The Normalised Triplets distance is similar between the Maximum Likelihood trees and the Bayesian consensus trees for the $M_{L}$ parameter (median Bhattacharrya Coefficient = @ %BC value for ML between both methods
; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}), however, for the two other parameters, $M_{F}$ and $M_{C}$, the Normalised Triplets distance is substantially higher for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = @ and @  %BC value for each parameter individually between both methods Tr
for $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

@@

\section{Acknowledgements}
Thanks to Gavin Thomas, Fr\'{e}d\'{e}ric Delsuc, Emmanuel Douzery, Trevor Hodkinson, Andrew Jackson, Gavin Thomas and April Wright for useful comments on our simulation protocol. Thanks to Paddy Doyle, Graziano D'Innocenzo and Sean McGrath for assistance with the computer cluster. Simulations used the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing and funded through grants from Science Foundation Ireland. This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).

@@

\bibliographystyle{sysbio}
\bibliography{References}

%---------------------------------------------
%
%       Supplementary, separated
%
%---------------------------------------------
